Cosmology Essay: A review of Dark Matter and Dark Energy

### 2017-06-21 cosmology/Essay/CosmoEssayHK.tex
2017Results

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.jpg}
\author{Howard Kinsman}
\title{Cosmology Essay: A review of Dark Matter and Dark Energy}
\begin{document}
\maketitle
\section{Introduction}
Many observations, and subsequent cosmological theories, have found that approximately 95\% of the mass and energy in the universe is dark and invisible, the true nature of which is
largely unknown. The matter component is known as dark matter (DM) and the energy component dark energy (DE) This is a brief review of the history and current status of 
DM and DE.
\section{Dark Matter}
DM has a surprisingly long and interesting history. In 1933, the renowned astrophysicist, Fritz Zwicky published a paper on the velocity dispersion of galaxies within the Coma cluster. He found
that the velocity dispersion of galaxies within the cluster was so high that in order for the system to remain stable there must be some ``missing mass'' i.e. mass which
is not detectable optically. Further research in this area continued but it wasn't until the 1970s when astronomers began accurately measuring the rotation curves of spiral galaxies,
with the discovery of HI 21cm radio emission, that interest in this area became of great importance. HI 21cm radio emission is produced by neutral hydrogen during a change of energy
state. It has a very precise wavelength of 21cm.
Rotation curves are a measure of the orbital velocity of the stars and gas in a galaxy plotted as a function of distance from the galactic centre. According to
Keplerian laws rotation curves would be expected to decline with increasing radius however it was found that they remain flat i.e. constant, all the way out to the largest radii observed, 
both optically and with 21 cm. According to Newtonian gravity this implies that there must be some ``hidden mass'' within the galaxies exerting a gravitational effect on the system resulting
in these flat rotation curves. This ``hidden mass'' was termed dark matter.
\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./rotationcurves.jpg}
\caption{Examples of flat galaxy rotation curves \citep{ruben}}
\label{fig:1}
\end{figure}

@@

Further evidence of DM, which has been described as ``direct empirical proof'' \citep{clowe}, has been found in observations of the Bullet cluster of galaxies. The Chandra X-ray observatory 
has detected the baryonic matter in the merger of two smaller clusters, whilst DM has been deduced from gravitational lensing, and they clearly show that are behaving differently.
At the collision point the baryonic matter has slowed down due to friction whilst the DM has passed through this point.
\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./bullet.jpg}
\caption{The bullet cluster: a) Optical view from Magellan telescope with plotted contours of mass distribution obtained from gravitational lensing.
b) Same contours but with Chandra X-ray data showing hot plasma. Most of the matter is lying in a different location to the hot plasma which has been slowed down by friction 
during the merger. \citep{clowe}}
\label{fig:2}
\end{figure}

@@

A new and exciting development by the European Space Agency, which will further probe DE, will be the launch of Euclid which is expected to be in 2020. Euclid aims to find out
the nature the acceleration of the expansion of the universe and DE. It will measure shapes and redshifts of galaxies and also explore the large-scale structure of the universe
by measuring the distribution of galaxy clusters. Euclid is a satellite equipped with a 1.2m telescope and three imaging and spectroscopic instruments working in optical and
near-infrared. It will be placed at the L2 Langrange point, a stable point in space, and is planned to operate for six years. The plan is cover 15000 square degrees of sky and
image one billion galaxies and take redshifts measurements of 100,000 galaxies \citep{amendola}. With these data Euclid will able to detail clustering of galaxies out to redshift
z=2 and out to redshift z=3 utilising weak lensing. Redshift measurements and weak lensing will also be accompanied by measuring correlations with the CMB, obtaining galaxy cluster 
information, strong lensing and possibly even obtaining luminosity distance through supernovae Type Ia measurements.

@@

\bibitem[Cohen-Tannoudji(2015)]{cohen}Cohen-Tannoudji, G., 2015, The dark universe and quantum vacuum, arXiv:1507.00460
\bibitem[ESA (2013)]{planck}Planck, ESA, http://www.esa.int/Our_Activities/Space_Science/Planck/Planck_reveals_an_almost_perfect_Universe, Retrieved 5-5-2017

>>>>

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listing}
\usepackage{url}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.jpg}
\author{Howard Kinsman}
\title{Cosmology Essay: A review of Dark Matter and Dark Energy}
\begin{document}
\maketitle
\section{Introduction}
Many observations, and subsequent cosmological theories, have found that approximately 95\% of the mass and energy in the universe is dark and invisible, the true nature of which is
largely unknown. The matter component is known as dark matter (DM) and the energy component dark energy (DE) This is a brief review of the history and current status of 
DM and DE.
\section{Dark Matter}
DM has a surprisingly long and interesting history. In 1933, the renowned astrophysicist, Fritz Zwicky published a paper on the velocity dispersion of galaxies within the Coma cluster. He found
that the velocity dispersion of galaxies within the cluster was so high that in order for the system to remain stable there must be some ``missing mass'' i.e. mass which
is not detectable optically. Further research in this area continued but it wasn't until the 1970s when astronomers began accurately measuring the rotation curves of spiral galaxies,
with the discovery of HI 21cm radio emission, that interest in this area became of great importance. HI 21cm radio emission is produced by neutral hydrogen during a change of energy
state. It has a very precise wavelength of 21cm.
Rotation curves are a measure of the orbital velocity of the stars and gas in a galaxy plotted as a function of distance from the galactic centre. According to
Keplerian laws rotation curves would be expected to decline with increasing radius however it was found that they remain flat i.e. constant, all the way out to the largest radii observed, 
both optically and with 21 cm. According to Newtonian gravity this implies that there must be some ``hidden mass'' within the galaxies exerting a gravitational effect on the system resulting
in these flat rotation curves. This ``hidden mass'' was termed dark matter.
\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./rotationcurves.jpg}
\caption{Examples of flat galaxy rotation curves \citep{ruben}}
\label{fig:1}
\end{figure}

@@

Further evidence of DM, which has been described as ``direct empirical proof'' \citep{clowe}, has been found in observations of the Bullet cluster of galaxies. The Chandra X-ray observatory 
has detected the baryonic matter in the merger of two smaller clusters, whilst DM has been deduced from gravitational lensing, and they clearly show that they are behaving differently.
At the collision point the baryonic matter has slowed down due to friction whilst the DM has passed through this point.
\begin{figure}[H]
\centering
\includegraphics[width=.9\textwidth]{./bullet.jpg}
\caption{The bullet cluster: a) Optical view from Magellan telescope with plotted contours of mass distribution obtained from gravitational lensing.
b) Same contours but with Chandra X-ray data showing hot plasma. Most of the matter is lying in a different location to the hot plasma which has been slowed down by friction 
during the merger. \citep{clowe}}
\label{fig:2}
\end{figure}

@@

A new and exciting development by the European Space Agency, which will further probe DE, will be the launch of Euclid which is expected to be in 2020. Euclid aims to find out
the nature the acceleration of the expansion of the universe and DE. It will measure shapes and redshifts of galaxies and also explore the large-scale structure of the universe
by measuring the distribution of galaxy clusters. Euclid is a satellite equipped with a 1.2m telescope and three imaging and spectroscopic instruments working in optical and
near-infrared. It will be placed at the L2 Langrange point, a stable point in space, and is planned to operate for six years. The plan is to cover 15000 square degrees of sky and
image one billion galaxies and take redshifts measurements of 100,000 galaxies \citep{amendola}. With these data Euclid will able to detail clustering of galaxies out to redshift
z=2 and out to redshift z=3 utilising weak lensing. Redshift measurements and weak lensing will also be accompanied by measuring correlations with the CMB, obtaining galaxy cluster 
information, strong lensing and possibly even obtaining luminosity distance through supernovae Type Ia measurements.

@@

\bibitem[Cohen-Tannoudji(2015)]{cohen}Cohen-Tannoudji, G., 2015, The dark universe and quantuxm vacuum, arXiv:1507.00460
\bibitem[ESA (2013)]{planck}Planck, ESA, \url{www.esa.int/Our_Activities/Space_Science/Planck/Planck_reveals_an_almost_perfect_Universe}, Retrieved 5-5-2017
